How kids code and how we know it:\\An exploratory study in the Scratch project repository

### 2016-04-20 Paper/paper.tex
added remark on init of vars
First version of research question 3 answers

\title{How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository }

@@

\subsubsection{Sprites}
Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{fig/ui.png}
		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right. The upper right corner shows the actual location of the sprite.}
		\label{fig:ui}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=2cm]{fig/bump.png}
		\caption{The script for the Piano sprite in Figure \ref{fig:ui}, causing a sound on receiving the `bump' signal.}
		\label{fig:bump}
	\end{center}
\end{figure}

\subsubsection{Events}
Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events. The canonical event is the `when Green Flag clicked', activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: `when Green Flag clicked' and `when right arrow key pressed', each with associated blocks.

\subsubsection{Signals}
Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the `when I receive' Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. The Piano reacts on that by making a sound, see Figure \ref{fig:bump}.

\subsubsection{Scripts}
Source code within sprites is organized in scripts: a script can start with an event (a Hat block\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}) which indicates when something should happen. A hat block can be followed by an arbitrary number of other blocks.

The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.

\subsubsection{Procedures}
Scratch users can define their own blocks, with a pop-up screen displayed in on the left hand side of Figure \ref{fig:twoBlocks}. As shown in this Figure, Scratch users can set the name of the sprite and define input parameters. The block in Figure \ref{fig:twoBlocks} uses two parameter, one textual (string1) and one numerical (number1). A user-created block is programmed using ordinary Scratch blocks as shown on the right hand side of Figure \ref{fig:twoBlocks}.

\subsubsection{Remixing}
\todo{maybe only useful with we will talk about remixing in the remainder?}
Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}. Shared Scratch programs can be `remixed' by other Scratch users, which means that a copy of this program is placed in the user's own project collection, and can be then further changed. The `remix tree' of projects is public, so users can track which users remix their programs, a bit similar forking in GitHub. Contrary to forking though, changes upstream cannot be integrated back into the original project.

\begin{figure}
	\begin{center}
		\includegraphics[width=8cm]{fig/twoBlocks.png}
		\caption{The user interface for creating and defining user-defined blocks \todo{could be nicer}}
		\label{fig:twoBlocks}
	\end{center}
\end{figure}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs. To answer our research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository. The dataset is described in Section \ref{dataset}. \todo{Below, we present how we approached each research question.}

@@

Kragle, as well as all scraped programs and our analysis files are available\footnote{\url{https://github.com/ScratchLover42/ICER-Data-Code}}.

@@

In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.

@@

The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default. Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage per project&4.80&0&0&0&3&2613\\
		LOC in Sprites per project&115.57&0&10&26&68&34613\\
		LOC in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Recursive loop statements per project&7.65&0&1&2&5&2503\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Histogram of the number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Histogram of the number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Histogram of the lines of code in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
	\caption{Histogram of the maximum code indentation level in the analyzed projects}
	\label{fig:size-indent}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks. For example, in Figure \ref{fig:ui} the cat sprite contains two scripts. We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. The maximum code indentation level is used as a code complexity metric. It was calculated using the JSON array depth? \todo{felienne, how did we measure this?}. This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

@@

The lines of code metric was further analyzed to understand code organization. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects. Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.

As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}. Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points. Most code is written in sprites. A small number of projects (7.7\%) use procedures, but they use them a lot and for more complex code. There exist surprisingly large and complex projects.
	\end{minipage}}

@@

The first method for abstraction that we investigate is the use of procedures. In the dataset we found 206,799 procedures in 17,979 projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\%\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
	\end{minipage}
	}

@@

\subsubsection{Code Clones Between Sprites}
With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common. In a sense, the Scratch users are not to blame here, as Scratch does not support the sharing of procedures between scripts, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

@@

\subsection{Threads to validity}
A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 programs. It could be the case that programming habits of Scratch users are changing over time. However, we counterbalanced that by using a very large dataset. Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}

>>>>

\title{How kids code and how we know it:\\An exploratory study in the Scratch project repository }

@@

\subsubsection{Sprites}
Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui}\footnote{\url{https://scratch.mit.edu/projects/97086781/}} consists of one sprite, the cat, which is Scratch's default sprite and logo. The code in the sprite will cause the cat to jump up, say ``hello'', and come back down, when the green flag is clicked, and to make the `meow' sound when the space bar is pressed.

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{fig/ui.png}
		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `motion' in the middle and the code associated with the sprite on the right. The upper right corner shows the actual location of the sprite.}
		\label{fig:ui}
	\end{center}
\end{figure}

\subsubsection{Signals}
\todo{uitleggen wat een signal is en evt ook al hoe het een kloon vervangt?}

\subsubsection{Events}
Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events. The canonical event is the `when Green Flag clicked', activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. In the Scratch code in Figure \ref{fig:ui} there are two events: `when Green Flag clicked' and `when space key pressed'.

\subsubsection{Scripts}
Source code within sprites is organized in scripts: a script always starts with an event, followed by a number of blocks. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously. For example, the code on the left of Figure \ref{fig:comparison} has five scripts associated with the `when Green Flag clicked' event.

\subsubsection{Remixing}
Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}. Shared Scratch programs can be `remixed' by other Scratch users, which means that a copy of this program is placed in the user's own project collection, and can be then further changed. The `remix tree' of projects is public, so users can track which users remix their programs, a bit similar forking in GitHub. Contrary to forking though, changes upstream cannot be integrated back into the original project.

\begin{figure}
	\begin{center}
		\includegraphics[width=6cm]{fig/makeBlockSmall.png}
		\caption{}
		\label{fig:makeBlock}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=6cm]{fig/useBlockSmall.png}
		\caption{}
		\label{fig:useBlock}
	\end{center}
\end{figure}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs. To answer our research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository. The dataset is described in Section \ref{dataset}. Below, we present how we approached each research question.

@@

Kragle, as well as all scraped programs and our analysis files are available via \todo{url} \todo{I do not think we need to describe the fields in the paper, but we need a good readme with the data (I can do that)}

@@

In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms Scratch projects in our dataset.

@@

The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default. Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage per project&4.80&0&0&0&3&2613\\
		LOC in Sprites per project&115.57&0&10&26&68&34613\\
		LOC in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Recursive loop statements per project&7.65&0&1&2&5&2503\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Histogram of the number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Histogram of the number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Histogram of the lines of code in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
	\caption{Histogram of the maximum code indentation level in the analyzed projects}
	\label{fig:size-indent}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks. For example, in Figure \ref{fig:ui} the cat sprite contains 2 scripts. We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. The maximum code indentation level is used as a code complexity metric. It was calculated using the JSON array depth? \todo{felienne, how did we measure this?}. This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

@@

The lines of code metric was further analyzed to understand code placement. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects. Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.

As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}. Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		\emph{RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points. Most code is written in sprites. Few projects (7.7\%) use procedures, but they use them a lot and for more complex code. There exist surprisingly large and complex projects.}
	\end{minipage}}

@@

As a first programming abstraction we investigated the use of procedures. In the dataset we found 206,799 procedures in 17,979 projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\%\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1052 (0.45\% of the total) projects.}
	\end{minipage}}

@@

\subsubsection{Code Clones Between Sprites}
With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common. In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of \todo{right word? procedures} between scripts, only within them. So in many cases there is no other way to share the functionality than making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.

@@
